### 2015-10-14 draft.tex
started wirting

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%Skeleton LaTeX file: double column format.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%REMEMBER THAT THERES IS AN EIGHT PAGE SIZE RESTRICTION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Sample file for ME Project Papers for Evaluation by Supervisor and Reader

\documentclass{article}

\usepackage{multicol}

\pagestyle{empty}
\setlength{\topmargin}{ 0.25in}
\setlength{\columnsep}{2.0pc}
\setlength{\headheight}{0.0in}
\setlength{\headsep}{0.0in}
\setlength{\oddsidemargin}{-.19in}
\setlength{\parindent}{1pc}
\textheight 8.75in
\textwidth 6.8in

\title{\large \bf Predicting Query Execution Time }
\author{Name}

\date{}

\begin{document}

	\maketitle
    \begin{center}
        Mid-term ME Project Report
    \end{center}
        \vskip 12pt
	\thispagestyle{empty}
	\bibliographystyle{unsrt}
		\begin{abstract}
		The ability to estimate the query execution time is central for a number of tasks in database system
		such as query scheduling, progress monitoring and costing during qury optimization. Recent work 
		has explored the use of statstical techniques in place of the manually constructed cost models used 
		in query optimization. Such techniques, which require as training data along with the 
		actual execution time, promises superior accuracies for they being able to account the for factors 
		such as hardware characterstics and bias in cardinality estimates. However, such techniques fail 
		to generalize i.e., produce poor estimates for queries that are not seen during the training.

		In this work, we propose and evaluate predictive modelling techniques that learn query 
		execution behaviour at a fine grained operator level. For each operator, we consider diffrent sets 
		of features and build diffrent models for them. Since there are only finitely many operators in 
		database, this approach is practical and will be able to estimate any query as its a composition of
		many operators. We evaluate our approches using TPC-H and TPC-DS workloads on PostgreSQL.

	\end{abstract}

	\hfill \\

	\begin{multicols}{2}
	\section{INTRODUCTION}
	Suspendisse fermentum, justo vitae suscipit convallis, nisl lorem commodo sem, ut ullamcorper neque mi in arcu. Quisque vitae tortor eget nisl faucibus tempus quis vel quam. Nullam quis facilisis sem. Mauris imperdiet nisl nec orci molestie ultricies. Maecenas tincidunt nisi sed tincidunt ultricies. Nunc malesuada ornare maximus. Nulla facilisi. Sed bibendum accumsan neque, quis pretium justo vestibulum at. Donec dapibus tortor nulla, et aliquam felis porta a. Sed suscipit quam massa, interdum accumsan eros tempus nec. Cras mattis, risus eu sagittis pretium, enim enim condimentum odio, in bibendum velit sem quis est. In euismod mi sit amet ante suscipit, ac suscipit nunc vulputate. Fusce at ipsum ac mauris lacinia fermentum.

	\section{2nd SECTION TITLE}
Pellentesque id leo eu massa rhoncus consectetur sit amet vel dui. Duis metus nibh, dignissim sit amet urna at, consequat semper libero. Fusce condimentum tortor mi, vel molestie sapien vehicula vitae. Integer ac ipsum sapien. Mauris lectus dolor, porta eget efficitur ac, varius a nulla. Fusce posuere dui ac nunc euismod, vitae convallis quam volutpat. Maecenas consequat malesuada porta. Nunc sit amet metus eget felis elementum porta. In eu vestibulum metus. Curabitur sed cursus purus.

%%%%%%%%%%%%%  This can be followed by several other sections

	\section{Conclusions and Future Work}
	Proin id elit nec erat pellentesque gravida ut a lectus. Proin rhoncus eu justo et aliquet. Praesent auctor augue quis magna dapibus imperdiet. Fusce vehicula vehicula aliquet. Fusce ac dolor in lorem tristique bibendum eget finibus turpis. In odio dui, mattis sed commodo non, bibendum et augue. Nam dignissim id metus ut bibendum. Nulla eu leo sed dui venenatis viverra. Donec et enim rutrum lorem aliquam bibendum ac ac arcu. Lorem ipsum dolor sit amet, consectetur adipiscing elit. In gravida enim vitae nisi tincidunt pulvinar. Maecenas hendrerit lacinia dui, et egestas turpis suscipit sed. Vestibulum eros orci, bibendum in diam in, suscipit aliquet ipsum. Nulla facilisi. Vestibulum vulputate, nulla ut ornare varius, eros leo elementum sem, at tempor ante massa vitae dolor.

    \bibliographystyle{abbrv}
    \bibliography{}

	\end{multicols}
\end{document}

### 2015-10-14 draft.tex
added introduction

@@
 \usepackage{multicol}
+\usepackage{graphicx}
 
@@
 		The ability to estimate the query execution time is central for a number of tasks in database system
-		such as query scheduling, progress monitoring and costing during qury optimization. Recent work 
-		has explored the use of statstical techniques in place of the manually constructed cost models used 
+		such as query scheduling, progress monitoring and costing during query optimization. Recent work 
+		has explored the use of statistical techniques in place of the manually constructed cost models used 
 		in query optimization. Such techniques, which require as training data along with the 
 		actual execution time, promises superior accuracies for they being able to account the for factors 
-		such as hardware characterstics and bias in cardinality estimates. However, such techniques fail 
+		such as hardware characteristics and bias in cardinality estimates. However, such techniques fail 
 		to generalize i.e., produce poor estimates for queries that are not seen during the training.
 
-		In this work, we propose and evaluate predictive modelling techniques that learn query 
-		execution behaviour at a fine grained operator level. For each operator, we consider diffrent sets 
-		of features and build diffrent models for them. Since there are only finitely many operators in 
+		In this work, we propose and evaluate predictive modeling techniques that learn query 
+		execution behavior at a fine grained operator level. For each operator, we consider different sets 
+		of features and build different models for them. Since there are only finitely many operators in 
 		database, this approach is practical and will be able to estimate any query as its a composition of
-		many operators. We evaluate our approches using TPC-H and TPC-DS workloads on PostgreSQL.
+		many operators. We evaluate our approaches using TPC-H and TPC-DS workloads on PostgreSQL.
 
@@
 	\section{INTRODUCTION}
-	Suspendisse fermentum, justo vitae suscipit convallis, nisl lorem commodo sem, ut ullamcorper neque mi in arcu.
-Quisque vitae tortor eget nisl faucibus tempus quis vel quam.
-Nullam quis facilisis sem.
-Mauris imperdiet nisl nec orci molestie ultricies.
-Maecenas tincidunt nisi sed tincidunt ultricies.
-Nunc malesuada ornare maximus.
-Nulla facilisi.
-Sed bibendum accumsan neque, quis pretium justo vestibulum at.
-Donec dapibus tortor nulla, et aliquam felis porta a.
-Sed suscipit quam massa, interdum accumsan eros tempus nec.
-Cras mattis, risus eu sagittis pretium, enim enim condimentum odio, in bibendum velit sem quis est.
-In euismod mi sit amet ante suscipit, ac suscipit nunc vulputate.
-Fusce at ipsum ac mauris lacinia fermentum.
+	Database systems can greatly benefit from accurate execution time predictions including:
+
+	\begin{itemize}
+	\item Query Optimizer: 
+	\item Query Scheduling:
+	\item Progress monitoring:
+	\end{itemize}
+
+	1D regression using linear, polynomial and RBF kernels.
+
+	\includegraphics[scale=.33]{figure_1.png}
 

### 2015-10-16 draft.tex
draft with more comments

@@
 	\end{abstract}		
-
-	\hfill \\
-
+	\hfill 
 	\begin{multicols}{2}
@@
 	Database systems can greatly benefit from accurate execution time predictions including: 
-
+	Explain the application briefly here.
 	\begin{itemize}
@@
 	\end{itemize}
-
-	1D regression using linear, polynomial and RBF kernels.
-
-	\includegraphics[scale=.33]{figure_1.png}
-
-	\section{2nd SECTION TITLE}
-Pellentesque id leo eu massa rhoncus consectetur sit amet vel dui.
-Duis metus nibh, dignissim sit amet urna at, consequat semper libero.
-Fusce condimentum tortor mi, vel molestie sapien vehicula vitae.
-Integer ac ipsum sapien.
-Mauris lectus dolor, porta eget efficitur ac, varius a nulla.
-Fusce posuere dui ac nunc euismod, vitae convallis quam volutpat.
-Maecenas consequat malesuada porta.
-Nunc sit amet metus eget felis elementum porta.
-In eu vestibulum metus.
-Curabitur sed cursus purus.
-
+	If possible plot a non-linear SVR for the data. will the optimizer's cost 
+	itself is enough 1D regression using linear, polynomial and RBF kernels
+	\includegraphics[scale=.33]{optcost.png}  
+	\section{Overview}
+	Elaborate what is the plan level approach. give  a intro model
+	 based techniques. i.e.,  OFF LINE TRAINING, emphasize one time.
+	1. How to obtain training data (explain analyze)
+	2. getting execution times at plan nodes
+	3. Individual model training 
+	4. Explain while testing and predict using Model more in next section
 %%%%%%%%%%%%%  This can be followed by several other sections
-
+	\section{Model selection and Training}
+	What model is considered good ? choice of model and kernel
+	If possible prove linear systems are not sufficient.
+	Training data ,specification. Does more training data helps?
+	 training time etc.. 
+	Essentially all operators need to be covered. generation QGEN tool. 
+	\section{Preliminary Experiments}
+	Give a background about Static and dynamic workload.  
+	do experiments at 2 stages
+	1. Act vs Act
+	2. Est vs Est(Bias towards estimated cardinalities)
+	For each type of above scenario, do testing for tpc-h queries 
+	(tpc-ds in future)
+	At different scales to see the model's ability to predict.
+	For midterm you should be having result for 1 as well as 10 GB.
+	For each above dataset, 
+	Results should convey L1 (MRE) as well as queries that are within 
+	[0-0.5] [0.5-1] [1-1.5] 
 	\section{Conclusions and Future Work}
-	Proin id elit nec erat pellentesque gravida ut a lectus.
-Proin rhoncus eu justo et aliquet.
-Praesent auctor augue quis magna dapibus imperdiet.
-Fusce vehicula vehicula aliquet.
-Fusce ac dolor in lorem tristique bibendum eget finibus turpis.
-In odio dui, mattis sed commodo non, bibendum et augue.
-Nam dignissim id metus ut bibendum.
-Nulla eu leo sed dui venenatis viverra.
-Donec et enim rutrum lorem aliquam bibendum ac ac arcu.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-In gravida enim vitae nisi tincidunt pulvinar.
-Maecenas hendrerit lacinia dui, et egestas turpis suscipit sed.
-Vestibulum eros orci, bibendum in diam in, suscipit aliquet ipsum.
-Nulla facilisi.
-Vestibulum vulputate, nulla ut ornare varius, eros leo elementum sem, at tempor ante massa vitae dolor.
+	Talk about generalization. 
+	operator specific features
+	capturing query interaction i.e., pipeline source of over-estimation
+	Optimizers provides sufficient information to detect a pipeline,
+	 can we use it to bind the operators and predict execution time 
+	 for a set of operators instead of a single operator.
 

### 2015-10-19 draft.tex
Done with most of the intro

@@
 	\thispagestyle{empty}
-	\bibliographystyle{unsrt}
+
 	\begin{abstract}
@@
 	Database systems can greatly benefit from accurate execution time predictions including: 
-	Explain the application briefly here.
 	\begin{itemize}
-	\item Query Optimizer: 
-	\item Query Scheduling:
-	\item Progress monitoring:
+%	\item Query Optimizer: With accurate estimates, optimizer can rely on this metric to select the best plan among the many available plans.
+	\item Admission control: Resource managers can use this metric to perform workload allocations such that the specific QoS are met \cite{activeSLA}.
+	\item Query Scheduling: Knowing the execution time is crucial in deadline and latency aware 			scheduling
+	\item Progress monitoring: Knowing the execution time of an incoming query can help avoid \textit{rogue queries} that are submitted in error and take an unreasonably long time to execute \cite{progress}.
 	\end{itemize}
-	If possible plot a non-linear SVR for the data. will the optimizer's cost 
-	itself is enough 1D regression using linear, polynomial and RBF kernels
-	\includegraphics[scale=.33]{optcost.png}  
-	\section{Overview}
+	Currently, execution time estimation is based on manually constructed
+	models, which are part of the query optimizer and typically use
+	combinations of weighted estimates of the number of tuples flowing
+	through operators, column widths, etc. Unfortunately, such
+	models often fail to capture several factors that affect the actual
+	resource consumption. For example, they may not include detailed
+	modeling of all of the various improvements made to database query
+	processing – such as nested loop optimizations \cite{ganapathi, adaptive} which localize
+	references in the inner subtree, and introduce “partially blocking”
+	batch sorts on the outer side, thereby increasing the memory
+	requirements and CPU time and reducing I/O compared to the traditional
+	iterator model. Similarly, they may not accurately reflect
+	specific hardware characteristics of the current production system
+	or the impact of cardinality estimation errors. Analytical cost models predominantly 
+	used by the current generation of query optimizers cannot
+	capture these interactions and complexity; in fact, they are not designed to do so. 
+	While they do a good job of comparing the costs of alternative query plans,
+	they are poor predictors of plan execution latency. 
+	Recent work \cite{ganapathi} showed this result for TPC-DS \cite{TPCDS}, and 
+	in this work we do the same for TPC-H \cite{TPCH} data and queries.
+
+	In this work, we utilize learning based models and prediction techniques which are promising reasonable accuracies in recent work \cite{ganapathi,MSR,ICDE2012}
+
+	\section{Background:Model Based Prediction}	
+COMEBACK:	Write about predictive stuff .. to give more intro. lets skip to imp parts now.
+
+\hfill\\
+
+	We use the term model to refer to any predictive function such as
+	Multiple Regression, Bayesian Nets, and Support Vector Machines.
+	Training a model involves using historical data sets to determine the
+	best model instance that explains the available data. For example,
+	fitting a function to a time series may yield a specific polynomial
+	instance that can be used to predict future values.
+	Model training (or building) requires selecting (i) the feature at-
+	tributes, a subset of all attributes in the data set, and (ii) a predic-
+	tion model, e.g., Linear Regression and Support Vector Machines
+	(SVMs), to be used for modeling. In general, we cannot know
+	which model type and feature set will produce the most accurate
+	model for a given data set without building and testing multiple
+	models. In some cases, a domain expert can manually specify the
+	feature attributes. In other cases, this step is trivial as the prediction
+	attribute(s) directly determine the feature attribute(s), e.g., in auto-
+	regressive models. Alternatively, feature attributes can be learned
+	automatically; however, given a set of n attributes, trying the power
+	set is prohibitively expensive if n is not small or training is expen-
+	sive [4, 3, 2] thereby requiring heuristic solutions.
+	Most approaches rank the candidate attributes (often based on their
+	correlation to the prediction attribute using metrics such as infor-
+	mation gain or correlation coefficients) and use this ranking to guide
+
+	a heuristic search [4] to identify most predictive attributes tested
+	over a disjoint test data set. In this paper, we will use a similar For-
+	ward Feature Selection algorithm based on linear correlation coef-
+	ficients [4]. This algorithm essentially performs a best-first search
+	in the model space. It starts with building models using small num-
+	ber of features and iteratively builds more complex and accurate
+	models by using more features. The features are considered with
+	respect to their correlation with the target/prediction attribute. The
+	training data set may be sampled to speed up the process.
+	While we use a feature selection algorithm for building accurate
+	models using relevant features, we do not consider building multi-
+	ple models of different types for solving the model selection prob-
+	lem. Instead in each one of our experiments we use a single type of
+	prediction model, either Linear Regression or SVMs, that performs
+	well.
+	Hypothesis testing and confidence interval estimations are two com-
+	mon techniques for determining predictive accuracy [2]. As men-
+	tioned, it is not possible to estimate a priori what model would be
+	most predictive for a given data set without training/testing it. One
+	form of hypothesis testing that is commonly used is K-Fold Cross
+	Validation (K-CV). K-CV divides the observed data up into k non-
+	overlapping partitions. One of the partitions is used as validation
+	data while the other k − 1 partitions are used to train the model
+	and to predict the data in the validation interval. In this study, we
+	will use cross-validation techniques to estimate the accuracy of our
+	prediction models.
+
+	\section{Overview of Proposed Approach}
+	\includegraphics[scale=0.3]{training.png}
+TODO :  
+1. One more flowchart for Testing. 
+2. Describe the concept of multiple models for a single operator. [MSR]
+3. How do we select the best one among them. (Mention that current experiments dont use this idea, only one a simple non-linear model is used for all models with 2 features)
+
 	Elaborate what is the plan level approach. give  a intro model
@@
 	4. Explain while testing and predict using Model more in next section
+
+	Mention assumptions, 
+	Pure in-memory.. lets solve this first.
+	For example, a sort merge
+
+join may require sorting its inputs. The sort operators are then responsible for
+handing over their result to the merge join via main memory. This means that
+the merge join may not require any I/O if the merge can be done purely in
+main memory
+
+Analytical approaches problems:
+
+counting the numbers of pages read is not
+sufficient as the discrepancy between random and sequential I/O is tremendous.
+
 %%%%%%%%%%%%%  This can be followed by several other sections
@@
 	\section{Preliminary Experiments}
+
+	If possible plot a non-linear SVR for the data. will the optimizer's cost 
+	itself is enough 1D regression using linear, polynomial and RBF kernels
+	\includegraphics[scale=.33]{optcost.png}
+
 	Give a background about Static and dynamic workload.  
@@
 
-    \bibliographystyle{abbrv}
-    \bibliography{}
+	\begin{thebibliography}{9}
+	\bibitem{opthet} 
+	Du, Weimin, Ravi Krishnamurthy, and Ming-Chien Shan. "Query optimization in a heterogeneous dbms." 		VLDB. Vol. 92. 1992.
+	\bibitem{activeSLA}
+	Xiong, Pengcheng, et al. "ActiveSLA: a profit-oriented admission control framework for database-as-a-		service providers." Proceedings of the 2nd ACM Symposium on Cloud Computing. ACM, 2011.	
+	\bibitem{progress}
+	Mishra, Chaitanya, and Nick Koudas. "The design of a query monitoring system." ACM Transactions on 		Database Systems (TODS) 34.1 (2009): 1.
 
+	\bibitem{ganapathi} 
+	A. Ganapathi, H. A. Kuno, U. Dayal, J. L. Wiener, A. Fox, M. I. Jordan,
+	and D. A. Patterson. Predicting multiple metrics for queries: Better
+	decisions enabled by machine learning. In ICDE, 2009.
+
+	\bibitem{MSR}
+	Li, Jiexing, et al. "Robust estimation of resource consumption for sql queries using statistical 			techniques." Proceedings of the VLDB Endowment 5.11 (2012): 1555-1566.
+
+	\bibitem{ICDE2012}
+	Akdere, Mert, et al. "Learning-based query performance modeling and prediction." Data Engineering 			(ICDE), 2012 IEEE 28th International Conference on. IEEE, 2012.
+
+     \bibitem{adaptive} 
+     S. Guirguis, M. A. Sharaf, P. K. Chrysanthis, A. Labrinidis, and
+	K. Pruhs. Adaptive scheduling of web transactions. In ICDE, 2009.
+
+	\bibitem{TPCH}
+	TPC-H benchmark specification, http://www.tpc.org/tpch/
+
+	\bibitem{TPCDS}
+	R. Othayoth and M. Poess, The making of tpc-ds, in VLDB
+	06: Proceedings of the 32nd international conference on Very
+	large data bases. VLDB Endowment, 2006, pp. 1049Ð1058
+
+	\end{thebibliography}
 	\end{multicols}

### 2015-10-21 draft.tex
Added experiments

@@
 \usepackage{graphicx}
+\usepackage{amsmath}
+\usepackage{framed}
 
@@
 	\section{Overview of Proposed Approach}
+
+	To-Do : improvise diagrams with creately.com \\
 	\includegraphics[scale=0.3]{training.png}
-TODO :  
-1. One more flowchart for Testing. 
-2. Describe the concept of multiple models for a single operator. [MSR]
-3. How do we select the best one among them. (Mention that current experiments dont use this idea, only one a simple non-linear model is used for all models with 2 features)
+	\includegraphics[scale=0.3]{testing.png}
 
-	Elaborate what is the plan level approach. give  a intro model
-	 based techniques. i.e.,  OFF LINE TRAINING, emphasize one time.
-	1. How to obtain training data (explain analyze)
-	2. getting execution times at plan nodes
-	3. Individual model training 
-	4. Explain while testing and predict using Model more in next section
+	In this section, we elaborate on the overview of the approach which consists of 2 phases. Like 
+	most other machine learning approaches, the first phase is an off-line training phase and the 
+	second is on-line testing phase.
+	During the training phase, we construct number of different 
+	models for each type of physical database operator(e.g.,  Sort-Merge Join, Nested Loop, Index Nested 		Loop,..).
+Each operator will be associated with so called 'Best-Fit' model which will be determined
+	based on the Operator model selection described in section 3.1. This model will be invoked for the
+	purpose of estimating the execution time for a certain operator.
+Since there are only finite number of 	operators and only a couple of physical implementations for them, the space consumption is not 				excessive.
+	\\
 
-	Mention assumptions, 
-	Pure in-memory.. lets solve this first.
-	For example, a sort merge
+	During the testing phase, we actually run the query to get the true cardinality estimates. While this
+	seems to contradict the whole estimation problem, right now this is required because the focus is on
+	solving the modeling problem given the actual cardinalities. The more practical scenario i.e., where
+	we we only the know the estimated cardinalities, is a much harder problem as now the models also need 
+	to be Robust to the errors in cardinality estimates. This is the natural follow-up work to do. Once 
+	after the execution of the query, we need to extract features depending on the operator and invoke 
+	the Best Fit model (which is already computed during training phase) to get an estimate. We sum up 
+	costs for all the operator present in the plan tree to finally give an estimate of execution time.
 
-join may require sorting its inputs. The sort operators are then responsible for
-handing over their result to the merge join via main memory. This means that
-the merge join may not require any I/O if the merge can be done purely in
-main memory
+TO-DO :  
+2. Describe the concept of multiple models for a single operator.
+3. How do we select the best one among them. (Mention that current experiments don't use this idea, only one a simple non-linear model is used for all models with 2 features)
 
@@
 
-counting the numbers of pages read is not
-sufficient as the discrepancy between random and sequential I/O is tremendous.
+counting the numbers of pages read is not sufficient as the discrepancy between random and sequential I/O is tremendous.
 
 %%%%%%%%%%%%%  This can be followed by several other sections
-	\section{Model selection and Training}
-	What model is considered good ? choice of model and kernel
-	If possible prove linear systems are not sufficient.
-	Training data ,specification. Does more training data helps?
-	 training time etc.. 
-	Essentially all operators need to be covered. generation QGEN tool. 
+	\section{Training and Model selection}
+	Before we describe the concept of multiple models for an operator, we shall present the motivation 
+	for doing so. While the statistical models can find the complex non-linear dependencies between 			features and the target, they cannot model the interaction among operators which might
+	affect the target value.
+
+	Consider the classic example of Sorting, where in the target variable (execution time) is 
+	proportional to $N_{l}$ log$N_{l}$. Unless this is used as a feature, the accuracy of the resulting 
+	model can be disappointing. Since we don't know the right function to be used beforehand, 
+	we explore the possible set of functions listed bellow and select the one with the 
+	minimum estimation error. We've used functions that are similar to the one's defined in 					\cite{robustIISc}.
+
+	\begin{itemize}
+	\item Linear: $a_{0} N_{l} + a_{1} N_{r}$ 
+	\item Quadratic: $a_{0} N_{l}N_{r}$ , $a_{0} N_{l}N_{l}$ , $a_{0} N_{r}N_{r}$ \ldots
+	\item Logarithmic: $a_{0} N_{l} log(N_{r}) , a_{0} N_{r} log(N_{l}) $ .. 
+	\item Exponential: $a_{0} N_{l} ^ {N_{r}}$  	
+	\end{itemize}
+
+	To illustrate the use of the functions in more details, we have plotted the curves with 
+	different possible functions for QuickSort as well as IndexNestedLoop operator. For the purpose
+	of generating curves for QuickSort we have used the following Query template
+
+	\begin{framed}
+	\begin{verbatim}
+	Select * from orders
+	where o_orderkey <= :varies 
+	order by o_totalprice
+	\end{verbatim}
+	\end{framed}
+
+	%We now show the observations for logarithmic and quadratic functions.
+	\includegraphics[scale=0.38]{Plots/quicksortnlogn.png}	
+	\includegraphics[scale=0.35]{Plots/quicksortnl2.png}
+
+	Unsurprisingly, logarithmic function fits better than the quadratic. 
+	While in some cases (Such as QuickSort) the appropriate function is obvious from SQL
+	query processing, it does not hold for all operators. 
+	For example, consider Nested Loop Join which 
+	can be modeled with quadratic functions in general but when the inner relation has an index 
+	the same model produces estimates estimates which are 
+	"off" by orders of magnitude. In the following example,
+	we take an exploratory approach to determine the best function for the Index Nested Loop operator.
+	As in the case of QuickSort we have used the following query template to generate the required
+	data.
+
+	\begin{framed}
+	\begin{verbatim}
+	select * from lineitem,orders
+	where l_orderkey = o_orderkey 
+		  and l_orderkey<= : varies;
+	\end{verbatim}
+	\end{framed}
+
+	\includegraphics[scale=0.37]{Plots/nil-nlnrlog2.png}	
+	\includegraphics[scale=0.37]{Plots/nil-nl+nr.png}	
+	\includegraphics[scale=0.37]{Plots/nil-nlnr.png}
+
+	Out of the above 3 functions, $N_{l} log(N_{r})$ fits data better than other functions. 	
+	\subsection{Best Fit Model}	
+	Best fit model $M_{O}$ for an Operator $O$ is the model which has the minimum estimation errors 
+	for the set of training queries. The estimation error is computed as follows,
+	\\
+	Model Estimation Error = \sum\limits_{i=1}^n \dfrac{\left| Actual_{i} - Estimated_{i} \right|}{n}
+
 	\section{Preliminary Experiments}
+	\subsection{Setup}
+	Our experimental study uses TPC-H decision support benchmark \cite{TPCH} on the top of 
+	PostgresSQL. The details are as follows,
+
+	\begin{itemize}
+	\item Database Management System : PostgreSQL 9.4
+	\item Data sets and workload: We created 1GB TPC-H database
+	according to the specification. The primary key indices as indicated
+	in the TPC-H specification were created for both databases. We have excluded Query 1,21  
+	To keep the training time under control, and Queries 15 as it creates a view which is not yet
+	supported in our work. This resulted in 19 out of 22 TPC-H queries being used for training. We've 
+	used TPC-H Qgen tool to generate 10 instances of each of the 19 queries, resulting in a total of 190 
+	queries.
+
+	\item Hardware: The queries were executed on a machine
+	with 8GB RAM running Ubuntu. buffer pool size was set to 1GB (25\% of the total RAM as the rule
+	of thumb). All queries were executed sequentially with cold start
+	(i.e., OS buffers flushed before the start of each query).
+
+	\item Statistical Models: As of now, we have limited our experiments to 2-features
+
+	TODO : Finish tomo. Talk with sir once.
+
+	TODO : 
+	0 . Optimzer vs Actual. 
+	1. Histogram of TPC-H errors.
+	2. table with distribution of MRE.
+	3. Performance operator wise. Box plot
+	4. Insights : which operators are hard to predict
+
+	\end{itemize}
+
+	\subsection{Prediction with Optimizer cost models}
 
@@
 	[0-0.5] [0.5-1] [1-1.5]
+
+%	Interpret tpc-h results by template wise, 
+%	To provide a sense of the magnitude of the estimation errors made
+%	by the optimizer, consider template-18, which is one of the templates 
+%	that exhibit the biggest error in operator-level prediction with
+%	actual/estimate model building. Instances of template-18 include 
+%	the following group by clause on table lineitem:
+%	group by l_orderkey having sum(l_quantity) > 314
+%	There are 15 million distinct l_orderkey values in lineitem
+%	(out of approximately 60 million tuples). The estimated number of
+%	groups satisfying sum(l_quantity) > 314 is 399521, whereas
+%	the actual number is 84. The PostgreSQL query optimizer com-
+%	putes this estimate using histograms (with 100 bins) for each col-
+%	umn based on the attribute independence assumption. The results
+%	are later fed into a Hash-Semi-Join, whose cost estimate is correspondingly
+%	very much off the mark.
+
 	\section{Conclusions and Future Work}
@@
 	\begin{thebibliography}{9}
-	\bibitem{opthet} 
-	Du, Weimin, Ravi Krishnamurthy, and Ming-Chien Shan. "Query optimization in a heterogeneous dbms." 		VLDB. Vol. 92. 1992.
-	\bibitem{activeSLA}
-	Xiong, Pengcheng, et al. "ActiveSLA: a profit-oriented admission control framework for database-as-a-		service providers." Proceedings of the 2nd ACM Symposium on Cloud Computing. ACM, 2011.	
-	\bibitem{progress}
-	Mishra, Chaitanya, and Nick Koudas. "The design of a query monitoring system." ACM Transactions on 		Database Systems (TODS) 34.1 (2009): 1.
-
 	\bibitem{ganapathi} 
@@
 
+	\bibitem{robustIISc}
+	Harish, D., Pooja N. Darera, and Jayant R. Haritsa. "Identifying robust plans through plan diagram 			reduction." Proceedings of the VLDB Endowment 1.1 (2008): 1124-1140.
+
 	\bibitem{TPCH}
@@
 
+		\bibitem{opthet} 
+	Du, Weimin, Ravi Krishnamurthy, and Ming-Chien Shan. "Query optimization in a heterogeneous dbms." 		VLDB. Vol. 92. 1992.
+	\bibitem{activeSLA}
+	Xiong, Pengcheng, et al. "ActiveSLA: a profit-oriented admission control framework for database-as-a-		service providers." Proceedings of the 2nd ACM Symposium on Cloud Computing. ACM, 2011.	
+	\bibitem{progress}
+	Mishra, Chaitanya, and Nick Koudas. "The design of a query monitoring system." ACM Transactions on 		Database Systems (TODS) 34.1 (2009): 1.
+
 	\end{thebibliography}

### 2015-10-25 draft.tex
Most of the midterm report done

@@
 	database, this approach is practical and will be able to estimate any query as its a composition of
-	many operators. We evaluate our approaches using TPC-H and TPC-DS workloads on PostgreSQL.
+	many operators. We evaluate our approaches using TPC-H workload on PostgreSQL.
 
@@
 
-	\section{Background:Model Based Prediction}	
-COMEBACK:	Write about predictive stuff .. to give more intro. lets skip to imp parts now.
-
-\hfill\\
-
+	\section{Background : Model Based Prediction}	
 	We use the term model to refer to any predictive function such as
-	Multiple Regression, Bayesian Nets, and Support Vector Machines.
+	Linear Regression, Least Squares and Support Vector Machines.
 	Training a model involves using historical data sets to determine the
@@
 	instance that can be used to predict future values.
-	Model training (or building) requires selecting (i) the feature at-
-	tributes, a subset of all attributes in the data set, and (ii) a predic-
+	Model training (or building) requires selecting (a) the feature at-
+	tributes, a subset of all attributes in the data set, and (b) a predic
 	tion model, e.g., Linear Regression and Support Vector Machines
-	(SVMs), to be used for modeling. In general, we cannot know
+	(SVMs), to be used for modeling. In general, we \textit{cannot} know
 	which model type and feature set will produce the most accurate
@@
 	models. In some cases, a domain expert can manually specify the
-	feature attributes. In other cases, this step is trivial as the prediction
-	attribute(s) directly determine the feature attribute(s), e.g., in auto-
-	regressive models. Alternatively, feature attributes can be learned
+	feature attributes. Alternatively, feature attributes can be learned
 	automatically; however, given a set of n attributes, trying the power
-	set is prohibitively expensive if n is not small or training is expen-
-	sive [4, 3, 2] thereby requiring heuristic solutions.
+	set is prohibitively expensive if n is not small or training is expensive 
+	thereby requiring heuristic solutions.
 	Most approaches rank the candidate attributes (often based on their
-	correlation to the prediction attribute using metrics such as infor-
-	mation gain or correlation coefficients) and use this ranking to guide
+	correlation to the prediction attribute using metrics such as information 
+	gain or correlation coefficients) and use this ranking to guide
+	a heuristic search \cite{datamining} to identify most predictive attributes tested
+	over a disjoint test data set. 
+%	In this paper, we will use a similar For-
+%	ward Feature Selection algorithm based on linear correlation coef-
+%	ficients [4]. This algorithm essentially performs a best-first search
+%	in the model space. It starts with building models using small num-
+%	ber of features and iteratively builds more complex and accurate
+%	models by using more features. The features are considered with
+%	respect to their correlation with the target/prediction attribute. The
+%	training data set may be sampled to speed up the process.
 
-	a heuristic search [4] to identify most predictive attributes tested
-	over a disjoint test data set. In this paper, we will use a similar For-
-	ward Feature Selection algorithm based on linear correlation coef-
-	ficients [4]. This algorithm essentially performs a best-first search
-	in the model space. It starts with building models using small num-
-	ber of features and iteratively builds more complex and accurate
-	models by using more features. The features are considered with
-	respect to their correlation with the target/prediction attribute. The
-	training data set may be sampled to speed up the process.
-	While we use a feature selection algorithm for building accurate
-	models using relevant features, we do not consider building multi-
-	ple models of different types for solving the model selection prob-
-	lem. Instead in each one of our experiments we use a single type of
-	prediction model, either Linear Regression or SVMs, that performs
-	well.
-	Hypothesis testing and confidence interval estimations are two com-
-	mon techniques for determining predictive accuracy [2]. As men-
-	tioned, it is not possible to estimate a priori what model would be
-	most predictive for a given data set without training/testing it. One
-	form of hypothesis testing that is commonly used is K-Fold Cross
-	Validation (K-CV). K-CV divides the observed data up into k non-
-	overlapping partitions. One of the partitions is used as validation
-	data while the other k − 1 partitions are used to train the model
-	and to predict the data in the validation interval. In this study, we
-	will use cross-validation techniques to estimate the accuracy of our
-	prediction models.
+	In this work we currently don't use any feature selection algorithm but rather rely on domain 
+knowledge of SQL Query processing.
+For the purpose of determining the best fit curve for the given set of features we consider building multiple models of different types.
+In each one of the models we use a single type of prediction model, either Linear Regression or SVMs, that performs well.
+
+	Hypothesis testing and confidence interval estimations are two common
+techniques for determining predictive accuracy \cite{forecasting}. As mentioned, 
+it is not possible to estimate a priori what model would be
+most predictive for a given data set without training/testing it. In this work, we are using
+disjoint training and testing datasets that are generated from qgen tool of TPC-H workload 		\cite{TPCH}. The mean relative error is used to estimate the accuracy of our prediction models.
 
 	\section{Overview of Proposed Approach}
+%	\begin{figure*}[h]
+%	  \includegraphics[width=\textwidth]{training.png}
+%	  \caption{Operator level approach to Execution time prediction}
+%	\end{figure*}
 
-	To-Do : improvise diagrams with creately.com \\
-	\includegraphics[scale=0.3]{training.png}
-	\includegraphics[scale=0.3]{testing.png}
+	\includegraphics[scale=0.4]{training.png}
 
@@
 Since there are only finite number of 	operators and only a couple of physical implementations for them, the space consumption is not 				excessive.
-	\\
+
+\hfill
 
@@
 	for the set of training queries. The estimation error is computed as follows,
-	\\
-	Model Estimation Error = \sum\limits_{i=1}^n \dfrac{\left| Actual_{i} - Estimated_{i} \right|}{n}
+
+	\begin{center}
+	$\dfrac{1}{n} \sum\limits_{i=1}^n \left| Actual_{i} - Estimated_{i} \right|$
+	\end{center}
 
@@
 	supported in our work. This resulted in 19 out of 22 TPC-H queries being used for training. We've 
-	used TPC-H Qgen tool to generate 10 instances of each of the 19 queries, resulting in a total of 190 
+	used TPC-H qgen tool to generate 10 instances of each of the 19 queries, resulting in a total of 190 
 	queries.
@@
 
-	\item Statistical Models: As of now, we have limited our experiments to 2-features
-
-	TODO : Finish tomo. Talk with sir once.
+	\item Statistical Models: As of now, we have limited our experiments to at-most 2 features (Input 			cardinalities $N_{l} , N_{r}$).
+For evaluating the best-fit model we have used linear regression 			models (available from python's sckit-learn \cite{sckit})
 
 	TODO : 
-	0 . Optimzer vs Actual. 
 	1. Histogram of TPC-H errors.
@@
 	\subsection{Prediction with Optimizer cost models}		
+	We start with the results showing that Optimizer's estimates are a poor indication of the real
+	execution time. For this purpose, we have taken 19 of 22 queries (To make a fair comparison with 
+	our approach we've excluded the 3 templates for the reason stated earlier). The actual execution time
+	and prediction times are computed using \textit{Explain Analyze} command of Postgres. The values 			collected are averaged over 5 runs.
+	\includegraphics[scale=0.45]{optcost.png}
 
-	If possible plot a non-linear SVR for the data. will the optimizer's cost 
-	itself is enough 1D regression using linear, polynomial and RBF kernels
-	\includegraphics[scale=.33]{optcost.png}
+	We've plotted the linear regression line resulting from fitting these points 
+	using linear least-squares regression 
+	(which can be seen as an error-minimizing mapping of the
+	optimizer-estimated cost (which is not measured in ms) to CPU-
+	time). As we can see, even for the mapping chosen to minimize
+	the overall error, there are significant differences between the estimated CPU cost 
+	and real CPU time for many queries. Similar
+	observations have been made in other works as well \cite{MSR,ICDE2012}
 
-	Give a background about Static and dynamic workload.  
-	do experiments at 2 stages
-	1. Act vs Act
-	2. Est vs Est(Bias towards estimated cardinalities)
-	For each type of above scenario, do testing for tpc-h queries 
-	(tpc-ds in future)
-	At different scales to see the model's ability to predict.
-	For midterm you should be having result for 1 as well as 10 GB.
-	For each above dataset, 
-	Results should convey L1 (MRE) as well as queries that are within 
-	[0-0.5] [0.5-1] [1-1.5]
+	Note that while the final estimate made by the optimizer may not be an accurate reflection of
+	real execution time, the estimates provided at a finer level (i.e., plan level, operator level etc.) 
+	are quite useful. In the next section, we show how we can use optimizer's estimates at operator level 
+	and produce an estimate for overall execution time.
 
-%	Interpret tpc-h results by template wise, 
-%	To provide a sense of the magnitude of the estimation errors made
-%	by the optimizer, consider template-18, which is one of the templates 
-%	that exhibit the biggest error in operator-level prediction with
-%	actual/estimate model building. Instances of template-18 include 
-%	the following group by clause on table lineitem:
-%	group by l_orderkey having sum(l_quantity) > 314
-%	There are 15 million distinct l_orderkey values in lineitem
-%	(out of approximately 60 million tuples). The estimated number of
-%	groups satisfying sum(l_quantity) > 314 is 399521, whereas
-%	the actual number is 84. The PostgreSQL query optimizer com-
-%	putes this estimate using histograms (with 100 bins) for each col-
-%	umn based on the attribute independence assumption. The results
-%	are later fed into a Hash-Semi-Join, whose cost estimate is correspondingly
-%	very much off the mark.
+	\subsection{Operator level modeling}
+	PostgreSQL query optimizer provides the wide range of information at multiple levels. For example, 
+	for the Hash operator, it provides the following information.
+
+	\begin{center}
+	\begin{tabular}{ |c|c| } 
+	 \hline
+		Node Type& Hash \\
+		Parent Relationship& Inner\\
+		Startup Cost& 4.45\\
+		Total Cost& 4.45\\
+		Plan Rows& 1\\
+		Plan Width& 117\\
+		Actual Startup Time& 0.017\\
+		Actual Total Time& 0.017\\
+		Actual Rows& 6\\
+		Actual Loops& 1\\
+		Hash Buckets& 1024\\
+		Hash Batches& 1\\
+		Original Hash Batches& 1\\
+		Peak Memory Usage& 1\\
+	 \hline
+	\end{tabular}
+	\end{center}
+
+	While all of these attributes may not be candidates for features (i.e may not impact execution time),	some do.
+We can use the knowledge of SQL Query processing to select relevant features and determine the Best Fit Model as described earlier(Section 4.1).
+In this set of experiments, we are limiting
+	ourselves to the feature 'Actual Rows' (i.e, Input cardinalities). In the case of binary operator nodes (such as
+	Nested Loop), we are considering both left and right input cardinalities. Please note that 
+	a) the proposed feature set should not be considered complete as it may not capture all the properties that impact execution time, we shall review these set of features from time to time as we make progress in this work.
+b) More number of features 	correspond to increased training time and since the computation of Best Fit Model is exponential to the number of features, we should select as few features as possible to make the computations tractable.
+
+	To ensure that the trained model is not over-fitting the data, we have generated training and testing queries do not contain identical instances of same query (i.e, Same query with same selectivities is not allowed).
+
+	\begin{center}
+	\begin{tabular}{ |c|c|c|c| } 
+	\hline
+	  Model & Minimum & Maximum & Mean\\ [0.5ex] 
+	 \hline
+	 Optimizer & 0.032  & 47.68 & 8.33 \\
+	 Our model & 0.003 & 2.61 & 0.53 \\
+	 \hline
+	\end{tabular}
+	\end{center}
+
+	\begin{figure*}[!t]
+	  \includegraphics[width=\textwidth]{tpch-1gbres.png}
+	  \caption{Operator-level modeling ,Errors by Template (1GB)}
+	\end{figure*}
+
+%	Give a background about Static and dynamic workload.  
+%	do experiments at 2 stages
+%	1. Act vs Act
+%	2. Est vs Est(Bias towards estimated cardinalities)
+%	For each type of above scenario, do testing for tpc-h queries 
+%	(tpc-ds in future)
+%	At different scales to see the model's ability to predict.
+%	For midterm you should be having result for 1 as well as 10 GB.
+%	For each above dataset, 
+%	Results should convey L1 (MRE) as well as queries that are within 
+%	[0-0.5] [0.5-1] [1-1.5]
+
+	\section {Related work}
 
 	\section{Conclusions and Future Work}
-	Talk about generalization. 
-	operator specific features
-	capturing query interaction i.e., pipeline source of over-estimation
-	Optimizers provides sufficient information to detect a pipeline,
-	 can we use it to bind the operators and predict execution time 
-	 for a set of operators instead of a single operator.
+	In this work, we have presented learning based approach at a finer granularity than whats proposed earlier in the literature.
+By learning at operator level we have shown that the model can generalize to the queries which are unseen during the training because of the fundamental property that execution time is upper bounded by execution time of individual operators present in the plan tree.
+	There remains a lot of work to be done.
+Currently feature set is limited, extending it will definitely improves accuracy and generality.
+However since the best-fit algorithm is exponential to number of features, it remains as a challenging work.
+%	We have assumed that we are getting true cardinalities, where in the practical scenario they will often have magnitude of errors.
+If we train the model with optimizers estimates instead of true cardinalities, the model to certain extent can have a bias towards estimation errors.
+On the other hand, we can have partial executions
+
+Also, interaction that happen within query makes accurate execution time prediction hard.
+For example, pipelining speeds up the query processing and is the primary source of over-estimation.
+Capturing these in the cost model is a challenging and has got the attention in recent years \cite{concurrent}.
+As mentioned earlier, we are not currently not addressing prediction in case of concurrent query execution.
+There is already some promising work in addressing this problem \cite{Ahmed,qshuffler}.
+It remains as an interesting future work to see how we can extend the features so that we can capture the execution time behavior under concurrent workloads.
 
@@
 
+	\bibitem{concurrent}
+	Wu, Wentao, et al. "Towards predicting query execution time for concurrent and dynamic database 			workloads." Proceedings of the VLDB Endowment 6.10 (2013): 925-936.
+
+	\bibitem{Ahmed}
+	Ahmad, Mumtaz, et al. "Modeling and exploiting query interactions in database systems." Proceedings of the 17th ACM conference on Information and knowledge management. ACM, 2008.
+
+	\bibitem{qshuffler}
+	Ahmad, Mumtaz, et al. "Qshuffler: Getting the query mix right." Data Engineering, 2008. ICDE 2008. IEEE 24th International Conference on. IEEE, 2008.
+
+	\bibitem{forecasting}
+	Makridakis, Spyros, Steven C. Wheelwright, and Rob J. Hyndman. Forecasting methods and applications. John Wiley, 2008.
+
 	\bibitem{TPCH}
@@
 	large data bases. VLDB Endowment, 2006, pp. 1049Ð1058
+
+	\bibitem{sckit}
+	Scikit-learn: Machine Learning in Python, Pedregosa et al., JMLR 12, pp. 2825-2830, 2011.
 

### 2015-10-25 draft.tex
Complete, pending revison

@@
 \title{\large \bf Predicting Query Execution Time }
-\author{Name}
+\author{Vamshi Pasunuru}
 
@@
 	TODO : 
-	1. Histogram of TPC-H errors.
-	2. table with distribution of MRE.
 	3. Performance operator wise. Box plot
@@
 
+	The benefits of modeling
+
 	\begin{figure*}[!t]
@@
 	\section {Related work}
+	Query-plan-level predictions have recently been studied \cite{ganapathi}. In \cite{ganapathi},
+	authors consider plan-level query performance prediction for the
+	following \textit{static} query workloads: the TPC-DS query benchmark
+	and a query workload obtained from a customer database. They
+	report that they can predict individual query execution times within
+	20\% of the actual time for 85\% of their test queries.	However their approach produces
+	poor estimates for queries which are \textit{different} from training queries. The ability to 
+	adapt to dynamic workload is crucial for the cost model to continuously adapt the changes in the
+	underlying databases and queries.
+
+	In \cite{ICDE2012, MSR} author's proposed an approach which learns at a finer granularity than 
+	at plan level. They use statistical models such as SVM and Linear regression to learn the behavior of 
+	every operator.
+They have shown by that learning at this level, the model can now recognize queries 		which can differ largely from the queries seen during training.
+They have only considered 14 out of 		22 templates and claimed that they were able to obtain 41\% i.e., an MRE of 0.41.
+We are different 			from their approach in 2 ways a) we consider learning at \textit{implementation level} i.e, for the operator \textit{sort}, to implement it optimizer has a choice among multiple sorting methods such as quick sort ,merge sort top-N heapsort and external sort etc.
+No previous approach distinguishes the implementation methods of an operator which is clearly incorrect because there's a substantial difference between resource consumptions of quick-sort and external sort as the former is CPU intensive (consumes relatively less time) and while the latter is IO intensive (consumes much more time).
+b) we consider multiple learning models for every implementation and select the one which fits the best.
+This approach is often used when the relation between features and target is unclear.
+
+	In \cite{analytical}, author's have done an analytical cost modeling where in they compute CPU and IO time individually by obtaining the number of operations performed and number of pages(sequential \& random) accessed respectively.
+They have also proposed a calibration phase where the cost model can account for the hardware changes which is absent in default Postgres model.
+In previous unpublished work by Phankuri, they have devised methods to compute random page cost which was left out in the work\cite{analytical}.
+While this kind of approach seems natural, there's inherent ambiguity in knowing the number of pages accessed.
+Even after knowing all this, there's no certain way we can compute IO time as cost model can never know whether a page is sequentially accessed or randomly accessed which impacts final cost computed heavily as there is a magnitude of difference in access times of sequential and random pages.
+	Finally, there has also been work on query progress indicators \cite{progress}. Query progress indicators provide estimations for the comple-
+	tion degrees of running queries. Such studies assume that the work
+	done by individual query operators are transparent, i.e., externally
+	visible. While these studies are also related to query execution performance,
+	they do not provide predictions for the execution time of
+	queries.
 
@@
 However since the best-fit algorithm is exponential to number of features, it remains as a challenging work.
+
 %	We have assumed that we are getting true cardinalities, where in the practical scenario they will often have magnitude of errors.
@@
 	Ahmad, Mumtaz, et al. "Modeling and exploiting query interactions in database systems." Proceedings of the 17th ACM conference on Information and knowledge management. ACM, 2008.
+	\bibitem{analytical}
+	Wu, Wentao, et al. "Predicting query execution time: Are optimizer cost models really unusable?." Data Engineering (ICDE), 2013 IEEE 29th International Conference on. IEEE, 2013.
+
+	\bibitem{progress}
+	Chaudhuri, S., Narasayya, V., and Ramamurthy, R.
+	Estimating progress of execution for SQL queries. In
+	Proceedings of the 2004 ACM SIGMOD international
+	Conference on Management of Data
 
@@
 	Xiong, Pengcheng, et al. "ActiveSLA: a profit-oriented admission control framework for database-as-a-		service providers." Proceedings of the 2nd ACM Symposium on Cloud Computing. ACM, 2011.
-	\bibitem{progress}
-	Mishra, Chaitanya, and Nick Koudas. "The design of a query monitoring system." ACM Transactions on 		Database Systems (TODS) 34.1 (2009): 1.
 

### 2015-10-25 draft.tex
Revision #1

@@
 	in query optimization. Such techniques, which require as training data along with the 
-	actual execution time, promises superior accuracies for they being able to account the for factors 
+	actual execution time, promises superior accuracies for they being able to account for the factors 
 	such as hardware characteristics and bias in cardinality estimates. However, such techniques fail 
@@
 	In this work, we propose and evaluate predictive modeling techniques that learn query 
-	execution behavior at a fine grained operator level. For each operator, we consider different sets 
-	of features and build different models for them. Since there are only finitely many operators in 
-	database, this approach is practical and will be able to estimate any query as its a composition of
-	many operators. We evaluate our approaches using TPC-H workload on PostgreSQL.
+	execution behavior at a fine grained operator level.
+For each \textit{physical implementation} of an 			operator, we consider different sets of features and build multiple models for them.
+Out of all those 		models, only the model that fits best is used later for prediciton.Since there are only finitely many 		operators in database, this approach is practical and will be able to estimate any query as its a 			composition of many operators.
+We evaluate our approaches using TPC-H workload on PostgreSQL.
 
@@
 	modeling of all of the various improvements made to database query
-	processing – such as nested loop optimizations \cite{ganapathi, adaptive} which localize
-	references in the inner subtree, and introduce “partially blocking”
-	batch sorts on the outer side, thereby increasing the memory
-	requirements and CPU time and reducing I/O compared to the traditional
-	iterator model. Similarly, they may not accurately reflect
+	processing – such as nested loop optimizations \cite{nestedloops}. 
+	Similarly, they may not accurately reflect
 	specific hardware characteristics of the current production system
@@
 	best model instance that explains the available data. For example,
-	fitting a function to a time series may yield a specific polynomial
+	fitting a function to a dataset may yield a specific linear/polynomial
 	instance that can be used to predict future values.
@@
 	models. In some cases, a domain expert can manually specify the
-	feature attributes. Alternatively, feature attributes can be learned
-	automatically; however, given a set of n attributes, trying the power
-	set is prohibitively expensive if n is not small or training is expensive 
-	thereby requiring heuristic solutions.
-	Most approaches rank the candidate attributes (often based on their
-	correlation to the prediction attribute using metrics such as information 
-	gain or correlation coefficients) and use this ranking to guide
-	a heuristic search \cite{datamining} to identify most predictive attributes tested
-	over a disjoint test data set. 
+	feature attributes. 
+%	Alternatively, feature attributes can be learned
+%	automatically; however, given a set of n attributes, trying the power
+%	set is prohibitively expensive if n is not small or training is expensive 
+%	thereby requiring heuristic solutions.
+%	Most approaches rank the candidate attributes (often based on their
+%	correlation to the prediction attribute using metrics such as information 
+%	gain or correlation coefficients) and use this ranking to guide
+%	a heuristic search \cite{datamining} to identify most predictive attributes tested
+%	over a disjoint test data set. 
 %	In this paper, we will use a similar For-
@@
 most predictive for a given data set without training/testing it. In this work, we are using
-disjoint training and testing datasets that are generated from qgen tool of TPC-H workload 		\cite{TPCH}. The mean relative error is used to estimate the accuracy of our prediction models.
+disjoint training and testing datasets that are generated from qgen tool of TPC-H workload 		\cite{TPCH}. Mean relative error(MRE) is used to estimate the accuracy of our prediction models.
+
+Throughout the discussion of this work, we discriminate operator by its implementation method i.e., For example, implementations of Aggregate operator Group Aggregate, Hash Aggregate and Unique are all treated as different operators.
 
@@
 	During the training phase, we construct number of different 
-	models for each type of physical database operator(e.g.,  Sort-Merge Join, Nested Loop, Index Nested 		Loop,..).
-Each operator will be associated with so called 'Best-Fit' model which will be determined
-	based on the Operator model selection described in section 3.1. This model will be invoked for the
-	purpose of estimating the execution time for a certain operator.
+	models for each type of database operator. Each operator will be associated with so called 'Best-Fit' 		model which will be determined 
+	based on the model selection described in next section.
+Later, while in testing, only this model will 		be invoked for the purpose of estimating the execution time of an operator.
 Since there are 			only finite number of operators and only a couple of physical implementations for them, the space 			consumption is not excessive.
@@
 	costs for all the operator present in the plan tree to finally give an estimate of execution time.
-
-TO-DO :  
-2. Describe the concept of multiple models for a single operator.
-3. How do we select the best one among them. (Mention that current experiments don't use this idea, only one a simple non-linear model is used for all models with 2 features)
-
-Analytical approaches problems:
-
-counting the numbers of pages read is not sufficient as the discrepancy between random and sequential I/O is tremendous.
 
@@
 They have also proposed a calibration phase where the cost model can account for the hardware changes which is absent in default Postgres model.
-In previous unpublished work by Phankuri, they have devised methods to compute random page cost which was left out in the work\cite{analytical}.
-While this kind of approach seems natural, there's inherent ambiguity in knowing the number of pages accessed.
-Even after knowing all this, there's no certain way we can compute IO time as cost model can never know whether a page is sequentially accessed or randomly accessed which impacts final cost computed heavily as there is a magnitude of difference in access times of sequential and random pages.
+In previous unpublished work by Phankuri, methods are provided to compute cost for reading random which was ambiguous the work\cite{analytical}.
+While this kind of approach seems natural, counting the numbers of pages read is not sufficient as the discrepancy between random and sequential I/O is tremendous.
+Unfortunately optimizer can never know whether a page is sequentially accessed or randomly accessed.
+
 	Finally, there has also been work on query progress indicators \cite{progress}. Query progress indicators provide estimations for the comple-
@@
 
+	\bibitem{nestedloops}
+	DeWitt, David J., Jeffrey F.
+Naughton, and Joseph Burger.
+"Nested loops revisited." Parallel and Distributed Information Systems, 1993., Proceedings of the Second International Conference on.
+IEEE, 1993.
+
 	\bibitem{TPCH}

### 2015-11-01 draft.tex
Midesem final report

@@
 \usepackage{framed}
+\usepackage{array}
+\usepackage{fancyhdr}
 
-\pagestyle{empty}
+% Turn on the style
+\pagestyle{fancy}
+\fancyhf{} % sets both header and footer to nothing
+\renewcommand{\headrulewidth}{0pt}
+
 \setlength{\topmargin}{ 0.25in}
 \setlength{\columnsep}{2.0pc}
-\setlength{\headheight}{0.0in}
-\setlength{\headsep}{0.0in}
+%\setlength{\headheight}{0.0in}
+%\setlength{\headsep}{0.0in}
 \setlength{\oddsidemargin}{-.19in}
@@
 \author{Vamshi Pasunuru}
+
+% Clear the header and footer
+\fancyhead{}
+\fancyfoot{}
+% Set the right side of the footer to be the page number
+\fancyfoot[R]{\thepage}
 
@@
 	\begin{abstract}
-	The ability to estimate the query execution time is central for a number of tasks in database system
-	such as query scheduling, progress monitoring and costing during query optimization. Recent work 
-	has explored the use of statistical techniques in place of the manually constructed cost models used 
-	in query optimization. Such techniques, which require as training data along with the 
-	actual execution time, promises superior accuracies for they being able to account for the factors 
-	such as hardware characteristics and bias in cardinality estimates. However, such techniques fail 
-	to generalize i.e., produce poor estimates for queries that are not seen during the training.
+	The ability to estimate the query execution time is crucial for a number of tasks in database system such as query scheduling, progress monitoring and costing during query optimization.
+Recent work has explored the use of statistical techniques in place of the manually constructed cost models used in query optimization.
+Such techniques, which require training data along with actual query execution time, promises superior accuracy since they can account for hardware characteristics.
+While these techniques which learn at plan level when applied in a static workload can give estimates very close to the actual execution time, they fail to generalize i.e., produce poor estimates for queries which are \textit{different} from the ones seen during training.
+This suggests that modeling at plan level might not be sufficient.
 
-	In this work, we propose and evaluate predictive modeling techniques that learn query 
-	execution behavior at a fine grained operator level.
+In this work, we propose and evaluate predictive modeling techniques that learn query execution behavior at a fine grained implementation level.
 For each \textit{physical implementation} of an operator, we consider different sets of features and build multiple models for them.
-Out of all those 		models, only the model that fits best is used later for prediciton.Since there are only finitely many 		operators in database, this approach is practical and will be able to estimate any query as its a 			composition of many operators.
+Out of all those models, only the model that fits \textit{best} is used later for prediction.
+Since there are only finitely many operators in database, this approach is practical and will be able to generalize as a query is simply a composition of many operators.
 We evaluate our approaches using TPC-H workload on PostgreSQL.
@@
 	models, which are part of the query optimizer and typically use
-	combinations of weighted estimates of the number of tuples flowing
-	through operators, column widths, etc. Unfortunately, such
+	manually constructed cost models which is a function of input cardinalities, 
+	column widths, etc. Unfortunately, such
 	models often fail to capture several factors that affect the actual
-	resource consumption. For example, they may not include detailed
-	modeling of all of the various improvements made to database query
-	processing – such as nested loop optimizations \cite{nestedloops}. 
+	resource consumption. For example, they cannot handle dependencies within query operators 
+	such as pipelining which has a significant impact on the actual execution time.
 	Similarly, they may not accurately reflect
@@
 	or the impact of cardinality estimation errors. Analytical cost models predominantly 
-	used by the current generation of query optimizers cannot
+	used by the current generation of query optimizer cannot
 	capture these interactions and complexity; in fact, they are not designed to do so. 
 	While they do a good job of comparing the costs of alternative query plans,
-	they are poor predictors of plan execution latency. 
-	Recent work \cite{ganapathi} showed this result for TPC-DS \cite{TPCDS}, and 
-	in this work we do the same for TPC-H \cite{TPCH} data and queries.
+	they are poor predictors of plan execution latency. To illustrate this, we have shown of plot of 		
+	Estimates of Optimizer vs Actual execution time by running TPC-H queries on 1GB in-memory Postgres 		
+	database.
 
@@
 	instance that can be used to predict future values.
-	Model training (or building) requires selecting (a) the feature at-
-	tributes, a subset of all attributes in the data set, and (b) a predic
-	tion model, e.g., Linear Regression and Support Vector Machines
+	Model training (or building) requires selecting (a) the feature attributes, 
+	a subset of all attributes in the data set, and (b) a prediction model, 
+	e.g., Linear Regression and Support Vector Machines
 	(SVMs), to be used for modeling. In general, we \textit{cannot} know
@@
 	feature attributes.
+
 %	Alternatively, feature attributes can be learned
@@
 most predictive for a given data set without training/testing it. In this work, we are using
-disjoint training and testing datasets that are generated from qgen tool of TPC-H workload 		\cite{TPCH}. Mean relative error(MRE) is used to estimate the accuracy of our prediction models.
+disjoint training and testing datasets that are generated from \textit{qgen} tool of TPC-H workload 		\cite{TPCH}.
+For quantifying the quality of predictions, we are using the metric called \textit{mean relative error}(MRE).
+	\begin{center}
+	MRE  = $\dfrac{1}{n} \sum\limits_{i=1}^n \dfrac{\left| Actual_{i} - Estimated_{i} \right|}{Actual_{i}}$
+	\end{center}
 
@@
 	\section{Overview of Proposed Approach}
-%	\begin{figure*}[h]
-%	  \includegraphics[width=\textwidth]{training.png}
-%	  \caption{Operator level approach to Execution time prediction}
-%	\end{figure*}
 
@@
 	models for each type of database operator. Each operator will be associated with so called 'Best-Fit' 		model which will be determined 
-	based on the model selection described in next section.
-Later, while in testing, only this model will 		be invoked for the purpose of estimating the execution time of an operator.
-Since there are 			only finite number of operators and only a couple of physical implementations for them, the space 			consumption is not excessive.
+	based on the model selection described in next section. Later, while in testing, only this model will 		be invoked for the purpose of estimating the execution time of an operator. 
+%	Since there are 			only finite number of operators and only a couple of physical implementations for them, the space 			consumption is not excessive.
 
@@
 
-	During the testing phase, we actually run the query to get the true cardinality estimates. While this
+	During the testing phase, we actually run the query to obtain true cardinalities. While this
 	seems to contradict the whole estimation problem, right now this is required because the focus is on
-	solving the modeling problem given the actual cardinalities. The more practical scenario i.e., where
-	we we only the know the estimated cardinalities, is a much harder problem as now the models also need 
-	to be Robust to the errors in cardinality estimates. This is the natural follow-up work to do. Once 
-	after the execution of the query, we need to extract features depending on the operator and invoke 
-	the Best Fit model (which is already computed during training phase) to get an estimate. We sum up 
-	costs for all the operator present in the plan tree to finally give an estimate of execution time.
+	solving the modeling problem given the true cardinalities. The more practical scenario i.e., where
+	we only the know the estimated cardinalities, is a much harder problem as now the models also 
+	need to be Robust to the errors in cardinality estimates. 
+	Once after the execution of the query, we need to extract features depending on the operator and 
+	invoke the Best Fit model (which is already computed during training phase) to get an estimate. We 
+	sum up costs for all the operators present in the plan tree to finally give an estimate of execution 
+	time.
 
 %%%%%%%%%%%%%  This can be followed by several other sections
-	\section{Training and Model selection}
-	Before we describe the concept of multiple models for an operator, we shall present the motivation 
-	for doing so. While the statistical models can find the complex non-linear dependencies between 			features and the target, they cannot model the interaction among operators which might
-	affect the target value.
+	\section{Training}
 
-	Consider the classic example of Sorting, where in the target variable (execution time) is 
-	proportional to $N_{l}$ log$N_{l}$. Unless this is used as a feature, the accuracy of the resulting 
-	model can be disappointing. Since we don't know the right function to be used beforehand, 
-	we explore the possible set of functions listed bellow and select the one with the 
-	minimum estimation error. We've used functions that are similar to the one's defined in 					\cite{robustIISc}.
+	\subsection{Features}
+	In this section, we describe set of possible features that we can use as input to operator level models.
+Features can be classified into global as well as local features.
+Global features are common to all operators while the local operators are specific to some operator(s).
 
+	\begin{center}
+	\begin{tabular}{|c|c|} 
+	\hline
+	  Name & Description \\ \hline
+	  $N_{l}$ & Number of input tuples in left child \\  \hline
+	  $N_{r}$ & Number of input tuples in right child \\  \hline
+	  $W_{l}$ & Tuple width of left child \\ \hline
+	  $W_{r}$ & Tuple width of right child \\ \hline
+	  COP & Names of child operators (Categorical) \\ \hline
+	  POP & Name of parent operator(Categorical) \\ \hline
+	  LOOPS & Number of loops \\ \hline
+	\end{tabular}
+	\\
+	\vspace{.5cm}
+	\textbf{Table 1: Global features}
+	\end{center}
+
+	\begin{center}
+	\begin{tabular}{|c|c|c|} 
+	\hline
+	  Description & Operator\\ \hline
+	  Levels of access in Index & IndexScan\\  \hline
+	  Sort Columns & Sort\\  \hline
+	  Columns involved in sort & Sort\\  \hline
+	  Columns involved in Hash & Hash\\  \hline
+	  Number of buckets & Hash\\  \hline
+	  Number of pages read & Scan\\  \hline
+	\end{tabular}
+	\\
+	\vspace{.5cm}
+	\textbf{Table 2: Local features}
+	\end{center}
+
+	Please note that a) the proposed feature set should not be considered complete as it may not capture all the properties that impact execution time, we shall review these set of features from time to time as we make progress in this work.
+b) More number of features 	correspond to increased training time and since the computation of Best Fit Model is exponential to the number of features, we should select as few features as possible to make the computations tractable.
+The task of computing which features are most important is not trivial and requires further analysis.
+
+	In this primarily work, we are considering only the 2 most important features $N_{l}$ and $N_{r}$ and the models were only built on these features.
+Including the full feature set will enable the model to generalize further.
+
+	\subsection{Model Selection}
+	The relation between features and target in our case can be determined either through a) Non linear regression model b) Linear regression model with transformed features.
+Through our experiments we realized that Non linear regression requires much more training data than linear regression model to obtain the same accuracy.
+Since from SQL processing knowledge we know the possible set of functions that can relate features to execution time we instead took the approach of linear regression.
+Here, we have used functions that are similar to the one's defined in \cite{robustIISc},
 	\begin{itemize}
@@
 
-	To illustrate the use of the functions in more details, we have plotted the curves with 
-	different possible functions for QuickSort as well as IndexNestedLoop operator. For the purpose
-	of generating curves for QuickSort we have used the following Query template
+	Since we don't know beforehand which function fits best without building 
+	and testing, we have to explore all the possible combinations among features. 
+	To illustrate the use of the functions in more details, we have plotted the curves
+	with different possible functions for Quick-Sort as well as NestedLoop-Index (NL-Index) operator. For the purpose of generating curves for Quick-Sort we have used the 
+	following Query template
 
@@
 
-	%We now show the observations for logarithmic and quadratic functions.
-	\includegraphics[scale=0.38]{Plots/quicksortnlogn.png}	
-	\includegraphics[scale=0.35]{Plots/quicksortnl2.png}
+%	\caption{Evaluating function $ N_{l} log(N_{l}$, this as we expected fits with very high accuracy }
+\begin{center}
+
+	\includegraphics[scale=0.37]{Plots/quicksortnlogn.png}
+	\textit{\small Evaluating $N_{l} log N_{l}$ for QuickSort}
+	\includegraphics[scale=0.37]{Plots/quicksortnl2.png}	
+	\textit{\small Evaluating $N_{l}*N_{l}$ for QuickSort}
+
+\end{center}
 
 	Unsurprisingly, logarithmic function fits better than the quadratic. 
-	While in some cases (Such as QuickSort) the appropriate function is obvious from SQL
+	While in some cases (such as Quick-Sort) the appropriate function is obvious from SQL
 	query processing, it does not hold for all operators. 
-	For example, consider Nested Loop Join which 
-	can be modeled with quadratic functions in general but when the inner relation has an index 
+	For example, consider Nested-Loop join which 
+	can be modeled with quadratic functions in general but when the inner relation has an 
+	index 
 	the same model produces estimates estimates which are 
-	"off" by orders of magnitude. In the following example,
-	we take an exploratory approach to determine the best function for the Index Nested Loop operator.
-	As in the case of QuickSort we have used the following query template to generate the required
+	"off" by orders of magnitude because now with the index on inner tables joins can be done much faster than before.
+Unless this is accounted separately in the learning model, it degrades the overall model for NestedLoop.
+This also explains why learning need to be done at implementation level and not at operator or plan level.
+
+Now we illustrate the same in the following example by taking exploratory approach to determine the best fit function for the NL-Index operator.
+	As in the case of Quick-Sort we have used the following query template to generate the required
 	data.
@@
 
+	\begin{center}
 	\includegraphics[scale=0.37]{Plots/nil-nlnrlog2.png}	
+	\textit{\small Evaluating $N_{l} log N_{r}$ for NL-Index}
 	\includegraphics[scale=0.37]{Plots/nil-nl+nr.png}	
+	\textit{\small Evaluating $N_{l} + N_{r}$ for NL-Index}
 	\includegraphics[scale=0.37]{Plots/nil-nlnr.png}	
+	\textit{\small Evaluating $N_{l} * N_{r}$ for Nl-Index}
+	\end{center}
 
-	Out of the above 3 functions, $N_{l} log(N_{r})$ fits data better than other functions. 	
+	Out of the above 3 functions, $N_{l} log(N_{r})$ fits data better than alternative functions which makes it the Best Fit Model for Nl-Index operator, the other models are discarded. 	
 	\subsection{Best Fit Model}	
-	Best fit model $M_{O}$ for an Operator $O$ is the model which has the minimum estimation errors 
+	Best fit model $M_{O}$ for an Operator $O$ is the model which has the minimum estimation error 
 	for the set of training queries. The estimation error is computed as follows,
@@
 	\begin{itemize}
-	\item Database Management System : PostgreSQL 9.4
+	\item Database Management System : Default PostgreSQL 9.4. Currently, we have not tuned 
+	any Postgres parameters. %which might improve its cost-model performance.
 	\item Data sets and workload: We created 1GB TPC-H database
-	according to the specification. The primary key indices as indicated
+	according to the specification. In this work, we are limiting ourselves to in-memory 
+	environment. The primary key indices as indicated
 	in the TPC-H specification were created for both databases. We have excluded Query 1,21  
-	To keep the training time under control, and Queries 15 as it creates a view which is not yet
-	supported in our work. This resulted in 19 out of 22 TPC-H queries being used for training. We've 
-	used TPC-H qgen tool to generate 10 instances of each of the 19 queries, resulting in a total of 190 
-	queries.
+	because the execution plans generated are not compliant with our current tree based 
+	model, and Queries 15 as it creates a view which is 
+	not yet	supported in our work. This resulted in 19 out of 22 TPC-H queries being used 
+	for training. We've used TPC-H qgen tool to generate 10 instances of each of the 19 
+	queries, resulting in a total of 190 queries.
 
 	\item Hardware: The queries were executed on a machine
-	with 8GB RAM running Ubuntu. buffer pool size was set to 1GB (25\% of the total RAM as the rule
-	of thumb). All queries were executed sequentially with cold start
-	(i.e., OS buffers flushed before the start of each query).
+	with 8GB RAM running Ubuntu. buffer pool size was set to 1GB. All queries were executed 
+	sequentially with cold start (i.e., OS buffers flushed before the start of each query).
 
-	\item Statistical Models: As of now, we have limited our experiments to at-most 2 features (Input 			cardinalities $N_{l} , N_{r}$).
-For evaluating the best-fit model we have used linear regression 			models (available from python's sckit-learn \cite{sckit})
-
-	TODO : 
-	3. Performance operator wise. Box plot
-	4. Insights : which operators are hard to predict
+	\item Statistical Models: For evaluating the best-fit model we have used linear regression and Support Vector Regression \cite{sckit}
 
@@
 	\subsection{Prediction with Optimizer cost models}		
-	We start with the results showing that Optimizer's estimates are a poor indication of the real
+	We start with the results showing that Optimize estimates are a poor indication of the real
 	execution time. For this purpose, we have taken 19 of 22 queries (To make a fair comparison with 
-	our approach we've excluded the 3 templates for the reason stated earlier). The actual execution time
-	and prediction times are computed using \textit{Explain Analyze} command of Postgres. The values 			collected are averaged over 5 runs.
+	our approach we've excluded the 3 templates for the reason stated earlier). The values collected are averaged over 5 runs.
+	\begin{center}
 	\includegraphics[scale=0.45]{optcost.png}
+	\\
+	\vspace{0.5cm}
+	\textbf{Figure 1 : Optimizer estimates can incur significant errors}
+	\end{center}
+	\setcounter{figure}{1}
 
+	Observe that lower left and upper left data points correspond to same
+	optimizer cost but differ significantly in the actual execution time.
 	We've plotted the linear regression line resulting from fitting these points 
@@
 	real execution time, the estimates provided at a finer level (i.e., plan level, operator level etc.) 
-	are quite useful. In the next section, we show how we can use optimizer's estimates at operator level 
+	are quite useful. In the next section, we show how we can use optimizer estimates at operator level 
 	and produce an estimate for overall execution time.
@@
 	\subsection{Operator level modeling}
-	PostgreSQL query optimizer provides the wide range of information at multiple levels. For example, 
-	for the Hash operator, it provides the following information.
+	\begin{figure*}[!t]
+	  \includegraphics[width=\textwidth]{tpch-1gbres.png}
+	  \caption{{Operator-level modeling ,Errors by Template (1GB)}}
+	\end{figure*}
+
+	To ensure that the trained model is not over-fitting the data, we have generated training and testing queries do not contain identical instances of same query (i.e, same query with same selectivities is not allowed).
 
 	\begin{center}
-	\begin{tabular}{ |c|c| } 
+	\begin{tabular}{|c|c|c|c|} 
 	\hline
-		Node Type& Hash \\
-		Parent Relationship& Inner\\
-		Startup Cost& 4.45\\
-		Total Cost& 4.45\\
-		Plan Rows& 1\\
-		Plan Width& 117\\
-		Actual Startup Time& 0.017\\
-		Actual Total Time& 0.017\\
-		Actual Rows& 6\\
-		Actual Loops& 1\\
-		Hash Buckets& 1024\\
-		Hash Batches& 1\\
-		Original Hash Batches& 1\\
-		Peak Memory Usage& 1\\
-	 \hline
+	  Model & Min RE & Max RE & Mean RE \\ 	 \hline
+	 Optimizer & 0.032  & 47.68 & 8.33 \\ \hline
+	 Our model & 0.003 & 2.61 & 0.53 \\ \hline
 	\end{tabular}
+	\\
+	\vspace{.5cm}
+	\textbf{Table 3: Relative error(RE) for TPC-H, 1GB}
 	\end{center}
 
-	While all of these attributes may not be candidates for features (i.e may not impact execution time),	some do.
-We can use the knowledge of SQL Query processing to select relevant features and determine the Best Fit Model as described earlier(Section 4.1).
-In this set of experiments, we are limiting
-	ourselves to the feature 'Actual Rows' (i.e, Input cardinalities). In the case of binary operator nodes (such as
-	Nested Loop), we are considering both left and right input cardinalities. Please note that 
-	a) the proposed feature set should not be considered complete as it may not capture all the properties that impact execution time, we shall review these set of features from time to time as we make progress in this work.
-b) More number of features 	correspond to increased training time and since the computation of Best Fit Model is exponential to the number of features, we should select as few features as possible to make the computations tractable.
-
-	To ensure that the trained model is not over-fitting the data, we have generated training and testing queries do not contain identical instances of same query (i.e, Same query with same selectivities is not allowed).
-
-	\begin{center}
-	\begin{tabular}{ |c|c|c|c| } 
-	\hline
-	  Model & Minimum & Maximum & Mean\\ [0.5ex] 
-	 \hline
-	 Optimizer & 0.032  & 47.68 & 8.33 \\
-	 Our model & 0.003 & 2.61 & 0.53 \\
-	 \hline
-	\end{tabular}
-	\end{center}
-
-	The benefits of modeling
-
-	\begin{figure*}[!t]
-	  \includegraphics[width=\textwidth]{tpch-1gbres.png}
-	  \caption{Operator-level modeling ,Errors by Template (1GB)}
-	\end{figure*}
+	The benefits of operator level modeling can be clearly
+	seen here. The optimizer cost model has a mean
+	MRE of 8.33 i.e., on an average it predicts within 800\%
+	actual execution time. Experiments on TPC-H queries
+	shows that optimizer cost estimate can degrade to as
+	much as 4700\%.
+	In Figure 2, we have shown the error obtained
+	for individual queries, our model produced estimates
+	that deviate by at-most 261\% and on-an average were
+	within 48\% of actual execution time. By using additional 
+	features we believe we can further bring this
+	down. To evaluate why certain queries are doing bad,
+	we dug into the plan tree and found that for the operators 
+	HashJoin, Materalize, BitmapHeapScan and
+	NestedLoop the model is producing estimates with
+	high MRE’s.
 
@@
 
-	In \cite{ICDE2012, MSR} author's proposed an approach which learns at a finer granularity than 
+	In \cite{ICDE2012, MSR} authors proposed an approach which learns at a finer granularity than 
 	at plan level. They use statistical models such as SVM and Linear regression to learn the behavior of 
@@
 We are different 			from their approach in 2 ways a) we consider learning at \textit{implementation level} i.e, for the operator \textit{sort}, to implement it optimizer has a choice among multiple sorting methods such as quick sort ,merge sort top-N heapsort and external sort etc.
-No previous approach distinguishes the implementation methods of an operator which is clearly incorrect because there's a substantial difference between resource consumptions of quick-sort and external sort as the former is CPU intensive (consumes relatively less time) and while the latter is IO intensive (consumes much more time).
+No previous learning based approach distinguishes the implementation methods of an operator which is clearly incorrect because there's a substantial difference between resource consumptions of quick-sort and external sort as the former is CPU intensive (consumes relatively less time) and while the latter is IO intensive (consumes much more time).
 b) we consider multiple learning models for every implementation and select the one which fits the best.
@@
 
-	In \cite{analytical}, author's have done an analytical cost modeling where in they compute CPU and IO time individually by obtaining the number of operations performed and number of pages(sequential \& random) accessed respectively.
+	In \cite{analytical}, authors have done an analytical cost modeling where in they compute CPU and IO time individually by obtaining the number of operations performed and number of pages(sequential \& random) accessed respectively.
 They have also proposed a calibration phase where the cost model can account for the hardware changes which is absent in default Postgres model.
-In previous unpublished work by Phankuri, methods are provided to compute cost for reading random which was ambiguous the work\cite{analytical}.
-While this kind of approach seems natural, counting the numbers of pages read is not sufficient as the discrepancy between random and sequential I/O is tremendous.
-Unfortunately optimizer can never know whether a page is sequentially accessed or randomly accessed.
+In previous report by Pankhuri, methods are provided to compute cost for reading random page which was ambiguous the work\cite{analytical}.
+While this kind of approach seems natural, interactions among query operators (such as pipelining)
 
-	Finally, there has also been work on query progress indicators \cite{progress}. Query progress indicators provide estimations for the comple-
-	tion degrees of running queries. Such studies assume that the work
+	Finally, there has also been work on query progress indicators \cite{progress}.
+Query progress indicators provide estimations for the completion degrees of running queries.
+Such studies assume that the work
 	done by individual query operators are transparent, i.e., externally
@@
 	In this work, we have presented learning based approach at a finer granularity than whats proposed earlier in the literature.
-By learning at operator level we have shown that the model can generalize to the queries which are unseen during the training because of the fundamental property that execution time is upper bounded by execution time of individual operators present in the plan tree.
+By learning at implementation level we have shown that the model can generalize to the queries which are unseen during the training.
 	There remains a lot of work to be done.
-Currently feature set is limited, extending it will definitely improves accuracy and generality.
-However since the best-fit algorithm is exponential to number of features, it remains as a challenging work.
+The feature set need to be extended to include operator specific features and parameters which can capture the effects of pipeline.
+The resulting model will be more accurate and also will be able to generalize to different queries and database scales.
 
@@
 
-	Also, interaction that happen within query makes accurate execution time prediction hard.
-For example, pipelining speeds up the query processing and is the primary source of over-estimation.
-Capturing these in the cost model is a challenging and has got the attention in recent years \cite{concurrent}.
-As mentioned earlier, we are not currently not addressing prediction in case of concurrent query execution.
-There is already some promising work in addressing this problem \cite{Ahmed,qshuffler}.
-It remains as an interesting future work to see how we can extend the features so that we can capture the execution time behavior under concurrent workloads.
+	\begin{thebibliography}{9}
+			\bibitem{analytical}
+	Wu, Wentao, et al. "Predicting query execution time: Are optimizer cost models really unusable?." ICDE 2013.
 
-	\begin{thebibliography}{9}
-	\bibitem{ganapathi} 
-	A. Ganapathi, H. A. Kuno, U. Dayal, J. L. Wiener, A. Fox, M. I. Jordan,
-	and D. A. Patterson. Predicting multiple metrics for queries: Better
-	decisions enabled by machine learning. In ICDE, 2009.
+	\bibitem{concurrent}
+	Wu, Wentao, et al. "Towards predicting query execution time for concurrent and dynamic database workloads." Proceedings of the VLDB Endowment (2013)
+
+	\bibitem{ICDE2012}
+	Akdere, Mert, et al. "Learning-based query performance modeling and prediction." ICDE 2012.
 
 	\bibitem{MSR}
-	Li, Jiexing, et al. "Robust estimation of resource consumption for sql queries using statistical 			techniques." Proceedings of the VLDB Endowment 5.11 (2012): 1555-1566.
+	Li, Jiexing, et al. "Robust estimation of resource consumption for sql queries using statistical techniques." Proceedings of the VLDB Endowment (2012)
 
-	\bibitem{ICDE2012}
-	Akdere, Mert, et al. "Learning-based query performance modeling and prediction." Data Engineering 			(ICDE), 2012 IEEE 28th International Conference on. IEEE, 2012.
+	\bibitem{sckit}
+	Scikit-learn: Machine Learning in Python, Pedregosa et al., JMLR 12, pp. 2825-2830, 2011.
+
+	\bibitem{activeSLA}
+	Xiong, Pengcheng, et al. "ActiveSLA: a profit-oriented admission control framework for database-as-a-service providers." Proceedings of the 2nd ACM Symposium on Cloud Computing. ACM, 2011.
+
+	\bibitem{ganapathi} 
+	Ganapathi, Archana, et al. "Predicting multiple metrics for queries: Better decisions enabled by machine learning." ICDE 2009.
 
     \bibitem{adaptive} 
-     S. Guirguis, M. A. Sharaf, P. K. Chrysanthis, A. Labrinidis, and
-	K. Pruhs. Adaptive scheduling of web transactions. In ICDE, 2009.
+     Guirguis, Shenoda, et al. "Adaptive scheduling of web transactions." ICDE 2009.
 
 	\bibitem{robustIISc}
-	Harish, D., Pooja N. Darera, and Jayant R. Haritsa. "Identifying robust plans through plan diagram 			reduction." Proceedings of the VLDB Endowment 1.1 (2008): 1124-1140.
+	Harish, D., Pooja N. Darera, and Jayant R. Haritsa. "Identifying robust plans through plan diagram reduction." Proceedings of the VLDB Endowment (2008)
 
-	\bibitem{concurrent}
-	Wu, Wentao, et al. "Towards predicting query execution time for concurrent and dynamic database 			workloads." Proceedings of the VLDB Endowment 6.10 (2013): 925-936.
-
-	\bibitem{Ahmed}
-	Ahmad, Mumtaz, et al. "Modeling and exploiting query interactions in database systems." Proceedings of the 17th ACM conference on Information and knowledge management. ACM, 2008.
-	\bibitem{analytical}
-	Wu, Wentao, et al. "Predicting query execution time: Are optimizer cost models really unusable?." Data Engineering (ICDE), 2013 IEEE 29th International Conference on. IEEE, 2013.
+	\bibitem{TPCDS}
+	R. Othayoth and M. Poess, The making of TPC-DS, Proceedings VLDB Endowment (2006).
 
 	\bibitem{progress}
-	Chaudhuri, S., Narasayya, V., and Ramamurthy, R.
-	Estimating progress of execution for SQL queries. In
-	Proceedings of the 2004 ACM SIGMOD international
-	Conference on Management of Data
-
-	\bibitem{qshuffler}
-	Ahmad, Mumtaz, et al. "Qshuffler: Getting the query mix right." Data Engineering, 2008. ICDE 2008. IEEE 24th International Conference on. IEEE, 2008.
-
-	\bibitem{forecasting}
-	Makridakis, Spyros, Steven C. Wheelwright, and Rob J. Hyndman. Forecasting methods and applications. John Wiley, 2008.
+	Chaudhuri, Surajit, Vivek Narasayya, and Ravishankar Ramamurthy.
+"Estimating progress of execution for SQL queries." Proceedings of the 2004 ACM SIGMOD international conference on Management of data.
+ACM, 2004.
 
@@
 
-	\bibitem{TPCDS}
-	R. Othayoth and M. Poess, The making of tpc-ds, in VLDB
-	06: Proceedings of the 32nd international conference on Very
-	large data bases. VLDB Endowment, 2006, pp. 1049Ð1058
-
-	\bibitem{sckit}
-	Scikit-learn: Machine Learning in Python, Pedregosa et al., JMLR 12, pp. 2825-2830, 2011.
-
-	\bibitem{opthet} 
-	Du, Weimin, Ravi Krishnamurthy, and Ming-Chien Shan. "Query optimization in a heterogeneous dbms." 		VLDB. Vol. 92. 1992.
-	\bibitem{activeSLA}
-	Xiong, Pengcheng, et al. "ActiveSLA: a profit-oriented admission control framework for database-as-a-		service providers." Proceedings of the 2nd ACM Symposium on Cloud Computing. ACM, 2011.
-
 	\end{thebibliography}

### 2015-11-24 draft.tex
syncing with lab's content

@@
 
-	In this work, we utilize learning based models and prediction techniques which are promising reasonable accuracies in recent work \cite{ganapathi,MSR,ICDE2012}
+	In this work, we utilize learning based models and prediction techniques which are promising reasonable accuracies in recent work \cite{ICDE2012,MSR,ganapathi}
 
@@
 	Hypothesis testing and confidence interval estimations are two common
-techniques for determining predictive accuracy \cite{forecasting}. As mentioned, 
+techniques for determining predictive accuracy. As mentioned, 
 it is not possible to estimate a priori what model would be
